---
title: "<strong>Advanced `R`</strong>"
author: "andrés castro araújo"
date: "`r Sys.Date()`"
output: 
  html_document: 
    code_folding: show
    theme: lumen
    toc: yes
    toc_float: 
      collapsed: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "", fig.align = "center")


# library(tidyverse)
# theme_set(theme_minimal(base_family = "Avenir", base_line_size = 0))
```


```{r results="asis", echo=FALSE, eval=FALSE}
cat("<style>
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 13px;
    border-left: 5px solid #eee;
}
</style>")
```

Taken from [Wickham](https://adv-r.hadley.nz)

****

## There's something about `R`

Knowing the difference between an object and its name will allow you to:

- More accurately predict the performance and memory usage of your code.

- Write faster code by avoiding accidental copies, a major source of slow code.

- Better understand R's functional programming tools.

We’ll use the `lobstr` package to dig into the internal representation of R objects.

```{r}
library(lobstr)
```

### Copy-on-modify semantics

Suppose we want to contain an object `x` that contains the values 1, 2, and 3.

```{r}
x <- c(1, 2, 3)
```

But this is _not_ an object with a name. R first creates a vector of values `c(1,2,3)` and then binds that object to a name `x`. 

>In other words, the object, or value, doesn’t have a name; it’s actually the name that has a value.

Thus, names are actually references to values. And if we now run `y <- x`, we don't get another copy of the value `c(1,2,3)`; instead, we get another binding to the existing object.

```{r, echo=FALSE, out.width="180px"}
knitr::include_graphics(
"https://d33wubrfki0l68.cloudfront.net/bdc72c04d3135f19fb3ab13731129eb84c9170af/f0ab9/diagrams/name-value/binding-2.png"
)
```

The `0x74b` refers to the object independently of its bindings. We can use `lobstr` to see that both `x` and `y` point to the same identifier. Note that these identifiers are long, and change every time you restart R.

```{r}
obj_addr(x)
y <- x
obj_addr(y)
```

Now suppose we modify `y`, but not `x`. 

```{r}
y[[2]] <- 100
```

R creates a new object `0xcd2`, which is basically of `0x74b` with one value change. Then `y` is rebound to that object.

```{r, echo=FALSE, out.width="180px"}
knitr::include_graphics(
"https://d33wubrfki0l68.cloudfront.net/ef9f480effa2f1d0e401d1f94218d0cf118433c0/b56e9/diagrams/name-value/binding-3.png"
)
```

>This behaviour is called __copy-on-modify.__ Understanding it will radically improve your intuition about the performance of R code.

This, for example, explains why loops perform better when we feed them a vector of adequate size.

```{r}
output <- vector("integer", length = 3)
for (i in 1:5) {
  output[[i]] <- i
  print(obj_addr(output)) ## remains the same while vector is length = 3
}
```

>Like vectors, lists use copy-on-modify behaviour; the original list is left unchanged, and R creates a modified copy. This, however, is a __shallow copy__: the list object and its bindings are copied, but the values pointed to by the bindings are not.











