---
title: "Python in R"
date: "`r Sys.Date()`"
output: 
  html_document: 
    theme: lumen
    toc: yes
    toc_float: 
      collapsed: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")

## Packages
library(reticulate)
library(tidyverse)
```

## Introduction

```{r}
### create a new environment 
# conda_create("r-NLP")
### install modules
# conda_install("r-NLP", c("scipy", "gensim", "networkx", "nltk", "numpy", "scikit-learn"))

# indicate that we want to use a specific condaenv
use_condaenv(condaenv = "r-NLP")
```

- Access to objects created within Python chunks from R using the `py` object (e.g. `py$x` would access an `x` variable created within Python from R).

- Access to objects created within R chunks from Python using the `r` object (e.g. `r.x` would access to `x` variable created within R from Python).

```{python}
## Python chunk
a = [1, 2, 5, "a"]
a

a.append("c")
a
```


```{r}
## R chunk
names(py$a) <- letters[1:5]
py$a    ## also can be used with stuff like dplyr::bind_rows() or tibble::enframe()
```

The translation between languages is somewhat awkward. Notice that we turned `py$a` into a _named list_, which in Python implies a `list` to `dictionary` convertion.

```{python}
## Python chunk
a
```

```{r}
## R chunk
## import NumPy (it will be automatically discovered in "r-reticulate")
np <- import("numpy", convert = TRUE)
np$random$random(10L)
```

```{python}
## Python chunk
import numpy as np
b = np.identity(3)
b
b.shape
```

```{r}
## R chunk
py$b
dim(py$b)
```

Just as we use `py$obj` to access Python objects from inside R chunks, we can use `r.obj` to access R objections from inside 

```{python}
## Python list comprehesion using an R object
[w for w in r.fruit if len(w) >= 12]
```

**Using Python functions**

```{python}
def flip(p):
    return np.random.random() < p
```

```{python}
## Python chunk
flip(0.5)
```

```{r}
## R chunk
replicate(10, py$flip(0.3))
```

```{python}
def all_pairs(nodes):
    for i, u in enumerate(nodes):
        for j, v in enumerate(nodes):
            if i>j:
                yield u, v
                
def random_pairs(nodes, p):
    for edge in all_pairs(nodes):
        if flip(p):
            yield edge
            
import networkx as nx
def make_random_graph(n, p):
    G = nx.Graph()
    nodes = range(n)
    G.add_nodes_from(nodes)
    G.add_edges_from(random_pairs(nodes, p))
    return G
```

```{python}
G = make_random_graph(n = 80, p = 0.01)
G = nx.adjacency_matrix(G).asformat("array")
```

```{r}
library(ggraph)
py$G %>% 
  igraph::graph_from_adjacency_matrix(mode = "undirected") %>% 
  ggraph(layout = "kk") +
  geom_edge_fan(alpha = 0.4) +
  geom_node_point(col = "steelblue1") +
  theme_graph()
```


## Other

There are many more things that __`reticulate`__ allows us to do (see [here](https://rstudio.github.io/reticulate/reference/index.html)).

Here are just a couple of interesting tidbits:

```{r}
## R chunk
keys <- c("a", "b", "c")
values <- 1:3
py_dict(keys, values)
py_dict(keys, values) %>% py_len()

## More complex objects
dict(list("andrés" = tuple(1, 2, 3)))
```

Creating Numpy arrays with `np_array()` and `array_reshape()` might come in handy too.

****

Finally, as of now RStudio's environment panel doesn't display the contents of the Python environment. There's no current analog to "magic commands" in Jupyter Notebooks, such as `%whos`, to do this, but these are a couple of ways around this problem:

- Solution 1: Use the tab-complete feature in RStudio when typing `py$...`

- Solution 2:

```{python}
[print(i) for i in locals().keys() if not i.endswith("__")]
```

- Solution 3: Wait until RStudio creates an addin.