---
title: "Shiny Tutorial"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: hide
    theme: paper
    toc: yes
    toc_float: 
      collapsed: no
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = "center",
                      fig.height = 4, fig.width = 5)
```

## Introduction

**The anatomy of a Shiny app**

1. **User interface**. It controls the layout and appearance of the app.

```
ui <- fluidPage()
```

2. **Server function.** Contains instructions needed to build the app.

```
server <- function(input, output) {}
```

3. **`shinyApp()`**. Creates the Shiny app object.

```
shinyApp(ui, server)
```

Finally. You can turn any HTML-oriented R Markdown document into a Shiny documents by adding **runtime: shiny** to the YAML metadata as a top-level field.

```
---
title: "Document"
output: html_document
runtime: shiny
---
```

### Hello World

```{r, message=FALSE, warning=FALSE}
library(shiny)
library(tidyverse)
theme_set(theme_minimal(base_family = "Palatino"))

make_pdf <- function(support, lambda) {
    ggplot(NULL, aes(x = support)) +
    stat_function(fun = dpois,
                  args = list(lambda = lambda),
                  geom = "bar", 
                  n = support[2] - support[1] + 1,
                  width = 0.4) +
    scale_x_continuous(
      breaks = seq(support[1], support[2], length.out = 11)
      ) +
  labs(x = "x", y = "density",
       title = "Poisson Probability Mass Function")
}

ui <- fluidPage(
  
  # Sidebar layout with a input and output definitions
  sidebarLayout(
    
    # Inputs
    sidebarPanel(
      
      # Select lambda
      sliderInput(inputId = "lambda",
                  label = "Lambda",
                  min = 0, 
                  max = 100,
                  value = 5,
                  step = 0.1),
      
      # Select Support
      sliderInput(inputId = "support", 
                  label = "Range of x-axis:",
                  min = 0, 
                  max = 200, 
                  value = c(0, 20), 
                  step = 10)
      
    ),
    
    # Output
    mainPanel(plotOutput(outputId = "pdf"))
  )
)

server <- function(input, output) {
  output$pdf <- renderPlot({
    make_pdf(input$support, input$lambda)
  }, height = 275)
}

shinyApp(ui, server)
```


## User Interface (`ui`)

We can understand the `ui` in terms of layers.

The first layer is the **`fluidPage()`** function, which consists on 

- `rows`, which makes elements appear in the same line; and 

- `columns`, which define how much horizontal space should each element occupy. 

The name "fluid page" indicates that all components are scaled automatically to fill all browser width. The default layout -`sidebarLayout()`- creates a **sidebar panel** and a wider **main panel**. 
    
In the "hello world" example we used `sliderInput()` to fill the sidebar, but we coul have also used other options like `selectInput()` (which you should always use with *named* vectors).

## Server function

The `server()` function defines the relationship between inputs and outputs. As such, it takes in two arguments: `input` and `output`, both of which are *named lists*. In other word, this function contains the instructions need to build the app.

In the "hello world" example, the input list contained `lambda` and `support`, and the output list contained the `pdf` object. This last object was defined inside the server function through the `renderPlot({})` function. 

As a general rule, each object to display is built with a "render" function. 

**Note.** The server function can also take in a third argument --`session`-- which is an environment that can be used to access information and functionality related to the session. But this involves a more advanced usage of Shiny.

## Inputs

- `selectInput(inputId, label, choices, selected, ...)`: Create a select list that can be used to choose a single or multiple items from a list of values.

- `sliderInput(inputId, label, min, max, value, ...)`: Constructs a slider widget to select a numeric value from a range.

- `actionButton(inputId, label, icon, ...)`: Creates an action button or link whose value is initially zero, and increments by one each time it is pressed.

- `checkboxInput(inputId, label, value, ...)`: Create a checkbox that can be used to specify logical values.

- `checkboxGroupInput()`: Create a group of checkboxes that can be used to toggle multiple choices independently. The server will receive the input as a character vector of the selected values.

- `dateInput(inputId, label, value, min, max, format, ...)`: Creates a text input which, when clicked on, brings up a calendar that the user can click on to select dates.

- `fileInput(inputId, label, ...)`: Create a file upload control that can be used to upload one or more files.

- Etc.

See [**here**](http://shiny.rstudio.com/gallery/widget-gallery.html) for more.

****

- The `req()` function ensures holds back the output from being calculated if the user input is missing or is expected to throw an error (see [**here**](https://shiny.rstudio.com/reference/shiny/latest/req.html)).

    Use req() where you might be inclined to write an if statement checking that an input value is available. It should come before the value that depends on the input is calculated.

## Outputs and Rendering

Note that each render function (like `renderPlot({})` or `renderTable({})`) is associated with its own output object (like `plotOutput` or `tableOutput`).

- `renderTable()`: Creates a reactive table that is suitable for assigning to an output slot.

- `renderImage()`: Renders a reactive image that is suitable for assigning to an output slot.

- `renderPlot()`: Renders a reactive plot that is suitable for assigning to an output slot.

- Etc.

## Reactive Programming

As we have seen, reactivity is established by linking an input with an output via a `render*()` function.

There are three types of objects in reactive programming.

- reactive `source` or a user input that comes (typically) through a browser interface;

- reactive `endpoint` or something that appears in the user's browser window (e.g. a plot, a table); and 

    One reactive source can be connected to multiple endpoints, and vice versa.

- reactive `conductor` or a component between source and endpoint.

We can create reactive expressions with the `reactive({})` function.

For example, we might want to create "reactive data frames" with dplyr verbs like "filer" and "select". In this example, the resulting data frame is a `conductor`. We do this to make our code more "readable", but also for efficiency purposes: reactive expressions are "lazy", they only get executed once, when the input changes. 

We can view all reactive expressions with `reactlog`. To do this we insert this line of code in our R script, before launching and pressing CTRL+F3 when the app is launched:

```
options(shiny.reactlog = TRUE)
```

**Stop-trigger-delay**

- `observeEvent(eventExpr, handlerExpr)` is used to perform an action in response to an event.

- `eventReactive(eventExpr, valueExpr)` is used to create a calculated value that only updates in response to an event (e.g. a user pressing an `actionButton`).

Both functions need to be explicitly told what triggers them.

- `isolate()` is used to stop a reaction. For example, if you isolate the title of a plot (e.g. `labs(title = isolate({...}))`), the plot will not change unless *other* inputs are also changed.

**Note**. Don't define reactive objects inside `render*()` functions.

## Customizing appearance

There are over 100 `tags` for constructing HTML documents provided in the Shiny package. These will translate simple R code into the appropriate HTML syntax.

For example:

```{r}
strong("this is bold")
em("this is italic")
p("This", em("is"), "a", strong("composite"), "thing")
```

These are the tag functions built into the Shiny package:

```{r}
names(tags)
```

See [**here**](https://shiny.rstudio.com/articles/tag-glossary.html) for more information.

The `wellPanel()` function is used to group similar kinds of inputs, in order to make it easier for the user to understand the app.

We use `tabPanel()` to create stackable panels, and `navlistPanel()` or `tabsetPanel()` to arrange them vertically or horizontally.