---
title: "<strong>Ten Great Ideas About Chance</strong>"
author: "andrés castro araújo"
date: "`r Sys.Date()`"
output: 
  html_document: 
    code_folding: show
    theme: lumen
    toc: yes
    toc_float:
      collapsed: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", comment = "")

library(tidyverse)
```             

```{css, echo=FALSE}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 14px;
    color: #828282;
    border-left: 10px solid #EEE;
}

body {
    font-size: 14px;
}
```

## The Weak Law of Large Numbers

>As an illustration [Bernoulli] considers an urn containing 3000 white pebbles and 2000 black pebbles and postulates independent draws with replacement of the pebble drawn. He asks whether one can find a number of draws so that it becomes "morally certain" that the ratio of white pebbles to black ones becomes approximately 3:2. He then chooses a high probability as moral certainty and establishes a number of draws sufficient to provide a positive answer.

```{r}
urn <- c(rep("white", 3e3), rep("black", 2e3)) %>% sample()
draws <- sample(urn, size = 1e4, replace = TRUE)
result <- cumsum(draws == "white") / cumsum(draws == "black")

tibble(result) %>% 
  mutate(draw = row_number()) %>% 
  ggplot(aes(x = draw, result)) + 
  geom_hline(yintercept = 1.5, color = "red") + 
  geom_line() 

result[length(result)]
```

```{r}
N <- 100
output <- vector("list", length = N)
for (i in seq_along(output)) {
  draws <- sample(urn, size = 1e4, replace = TRUE)
  output[[i]] <- cumsum(draws == "white") / cumsum(draws == "black")
}

names(output) <- paste0("s", 1:N)
df <- as_tibble(output)

df %>% 
  mutate(draw = row_number()) %>% 
  pivot_longer(-draw) %>% 
  ggplot(aes(draw, value, group = name)) + 
  geom_line(alpha = 0.1) + 
  coord_cartesian(ylim = c(1, 2))
```



BIRTHDAY AND COINCIDENCES


